---
title: 'PUBH 7485 Homework 4'
output: 
  github_document:
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(mediation)
library(ranger)
library(rpart)
library(rpart.plot)

options(scipen = 999)
options(digits = 3)
```

```{r warning = FALSE, message = FALSE}
# Read in data
load(here("data", "OPT_Study_Mediation_PUBH7485_8485_v2.Rdata"))
```

## Impute
```{r warning = FALSE, message = FALSE}
# Impute continuous data
# Calculate medians
median_bmi <- median(opt_mediation$BMI, na.rm = TRUE)
median_cigday <- median(opt_mediation$BL.Cig.Day, na.rm = TRUE)
median_drinksday <- median(opt_mediation$BL.Drks.Day, na.rm = TRUE)
median_prevpreg <- median(opt_mediation$N.prev.preg, na.rm = TRUE)
median_livingkids <- median(opt_mediation$N.living.kids, na.rm = TRUE)
median_birthweight <- median(opt_mediation$Birthweight, na.rm = TRUE)
median_V5..BOP <- median(opt_mediation$V5..BOP, na.rm = TRUE)

# Create new imputed columns
opt_mediation$bmi_imputed <- ifelse(is.na(opt_mediation$BMI), median_bmi, opt_mediation$BMI)
opt_mediation$cigday_imputed <- ifelse(is.na(opt_mediation$BL.Cig.Day), median_cigday,
  opt_mediation$BL.Cig.Day
)
opt_mediation$drinksday_imputed <- ifelse(is.na(opt_mediation$BL.Drks.Day),
  median_drinksday, opt_mediation$BL.Drks.Day
)
opt_mediation$prevpreg_imputed <- ifelse(is.na(opt_mediation$N.prev.preg),
  median_prevpreg, opt_mediation$N.prev.preg
)
opt_mediation$livingkids_imputed <- ifelse(is.na(opt_mediation$N.living.kids),
  median_livingkids, opt_mediation$N.living.kids
)
opt_mediation$birthweight_imputed <- ifelse(is.na(opt_mediation$Birthweight),
  median_birthweight, opt_mediation$Birthweight
)
opt_mediation$V5..BOP_imputed <- ifelse(is.na(opt_mediation$V5..BOP),
  median_V5..BOP, opt_mediation$V5..BOP
)
opt_mediation <- opt_mediation %>%
  fill(Any.live.ptb.sb.sp.ab.in.ab, .direction = "downup")

# Impute categorical data
# Create new imputed columns
opt_mediation$hisp_imputed <- ifelse(is.na(opt_mediation$Hisp), "Yes", opt_mediation$Hisp)
opt_mediation$diabtype_imputed <- ifelse(is.na(opt_mediation$BL.Diab.Type), "Type II",
  opt_mediation$BL.Diab.Type
)
opt_mediation$tobuse_imputed <- ifelse(is.na(opt_mediation$Use.Tob), "No", opt_mediation$Use.Tob)
opt_mediation$alcuse_imputed <- ifelse(is.na(opt_mediation$Use.Alc), "No", opt_mediation$Use.Alc)
opt_mediation$addiction_imputed <- ifelse(is.na(opt_mediation$Drug.Add), "No", opt_mediation$Drug.Add)



# Create change in proportion of sites that bleed variable
opt_mediation <- opt_mediation %>%
  mutate(BOP_change = (V5..BOP_imputed - BL..BOP))

# Create numeric group
opt_mediation <- opt_mediation %>%
  mutate(group_num = fct_recode(Group,
    "1" = "T",
    "0" = "C"
  )) %>%
  mutate(group_num = as.numeric(as.character(group_num)))

# Find median of mediator
median(opt_mediation$BOP_change)
# -7

```


## Question 1: Mediation analysis
### a. Mediator model 

```{r warning = FALSE, message = FALSE}
mediator <- lm(BOP_change ~ Group + White + Nat.Am + Diabetes + cigday_imputed + Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_mediation)
round(summary(mediator)$coefficients, digits = 3)
```

###b. Outcome model with mediator

```{r warning = FALSE, message = FALSE}
outcome_bw <- lm(birthweight_imputed ~ Group + BOP_change + Group:BOP_change + White + Nat.Am + Diabetes + cigday_imputed + Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_mediation)
round(summary(outcome_bw)$coefficients, digits = 3)

outcome_preg <- glm(Preg.ended...37.wk ~ Group + BOP_change + Group:BOP_change + White + Nat.Am + Diabetes + cigday_imputed + Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_mediation, family = binomial)
round(summary(outcome_preg)$coefficients, digits = 3)

```


###c. Calculating CDE and NIE

#### Birthweight
```{r warning = FALSE, message = FALSE}
# 7 is the treatment*mediator interaction (13 in mine)
# 2 is the treatment parameter
# 3 is the mediator parameter
# Note: NDE + NIE should be close to ATE
# CDE
CDE <- outcome_bw$coefficients[2] + (outcome_bw$coefficients[13]*-7)
CDE

### Add mean or median of the mediator value...

# NIE
opt_mediation0 <- opt_mediation
opt_mediation0$group_num <- 0
NIE <- outcome_bw$coefficients[3] * mediator$coefficients[2] +
  outcome_bw$coefficients[13] * mediator$coefficients[2]
NIE

# Note - the group has to be numeric
```


```{r warning = FALSE, message = FALSE}
# Bootstrap for SE and CI
n <- nrow(opt_mediation)
CDE.boot <- NULL
NIE.boot <- NULL
B <- 100
set.seed((2123))
for (i in 1:B) {
  opt_mediation.boot <- opt_mediation[sample(1:n, n, replace = TRUE), ]
  outcome.boot <- lm(birthweight_imputed ~ Group + BOP_change + Group:BOP_change + White + Nat.Am + Diabetes + cigday_imputed + Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_mediation.boot)
  mediator.boot <- lm(BOP_change ~ Group + White + Nat.Am + Diabetes + cigday_imputed + Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_mediation.boot)
  opt_mediation0.boot <- opt_mediation.boot
  opt_mediation0.boot$group_num <- 0
  CDE.boot <- c(
    CDE.boot,
    outcome.boot$coefficients[2] + (outcome.boot$coefficients[13]*-7)
  )
  NIE.boot <- c(
    NIE.boot,
     outcome.boot$coefficients[3] * mediator.boot$coefficients[2] + 
      outcome.boot$coefficients[13] * mediator.boot$coefficients[2]
  )
}
```


```{r warning = FALSE, message = FALSE}
# Results
CDE
SE_CDE <- sd(CDE.boot)
SE_CDE
CI_CDE <- CDE + c(-1, 1)*qnorm(0.975)*SE_CDE
CI_CDE


NIE
SE_NIE <- sd(NIE.boot) 
SE_NIE
CI_NIE <- NIE + c(-1, 1)*qnorm(0.975)*SE_NIE
CI_NIE
```

#### Pregnancy ended
```{r warning = FALSE, message = FALSE}
# 7 is the treatment*mediator interaction (13 in mine)
# 2 is the treatment parameter
# 3 is the mediator parameter
# if treatment*covariate interactions: add an additional term which is the expected value of the covariates

# Note: NDE + NIE should be close to ATE
# CDE
CDE <- outcome_preg$coefficients[2] + (outcome_preg$coefficients[13]*-7)
CDE

### Add mean or median of the mediator value...

# NIE
opt_mediation0 <- opt_mediation
opt_mediation0$group_num <- 0
NIE <- outcome_preg$coefficients[3] * mediator$coefficients[2] +
  outcome_preg$coefficients[13] * mediator$coefficients[2]
NIE

# Note - the group has to be numeric
```


```{r warning = FALSE, message = FALSE}
# Bootstrap for SE and CI
n <- nrow(opt_mediation)
CDE.boot <- NULL
NIE.boot <- NULL
B <- 100
set.seed((2123))
for (i in 1:B) {
  opt_mediation.boot <- opt_mediation[sample(1:n, n, replace = TRUE), ]
  outcome.boot <- glm(Preg.ended...37.wk ~ Group + BOP_change + Group:BOP_change + White + Nat.Am + Diabetes + cigday_imputed + Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_mediation.boot, family = binomial())
  mediator.boot <- lm(BOP_change ~ Group + White + Nat.Am + Diabetes + cigday_imputed + Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_mediation.boot)
  opt_mediation0.boot <- opt_mediation.boot
  opt_mediation0.boot$group_num <- 0
  CDE.boot <- c(
    CDE.boot,
    outcome.boot$coefficients[2] + (outcome.boot$coefficients[13]*-7)
  )
  NIE.boot <- c(
    NIE.boot,
     outcome.boot$coefficients[3] * mediator.boot$coefficients[2] + 
      outcome.boot$coefficients[13] * mediator.boot$coefficients[2]
  )
}
```


```{r warning = FALSE, message = FALSE}
# Results
CDE
SE_CDE <- sd(CDE.boot)
SE_CDE
CI_CDE <- CDE + c(-1, 1)*qnorm(0.975)*SE_CDE
CI_CDE


NIE
SE_NIE <- sd(NIE.boot) 
SE_NIE
CI_NIE <- NIE + c(-1, 1)*qnorm(0.975)*SE_NIE
CI_NIE
```



## Question 2: 

Fit flexible regression models for the birthweight and whether or not the pregnancy ended before 37 weeks outcomes in the OPT study, and find the estimated optimal treatment rule. 

```{r warning = FALSE, message = FALSE}
# Regression Output - birthweight
m1 <- glm(birthweight_imputed ~ Group*(Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I), data = opt_mediation, family = "gaussian")
round(summary(m1)$coefficients, digits = 3)

# Regression Output - pregnancy 
m3 <- glm(Preg.ended...37.wk ~ Group*(Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I), data = opt_mediation, family = "binomial")
round(summary(m3)$coefficients, digits = 3)
```

Please summarize the proportion who would benefit from treatment.

```{r warning = FALSE, message = FALSE}
# Determine Optimal Treatment Assignment for Each Subject
# Note: does not work well with the group_num variable - must be a factor?
# Birth weight
data_trt <- data_ctr <- opt_mediation
data_trt$Group = "T"
data_ctr$Group = "C"
pred1 <- predict(m1, newdata = data_trt, type = "response")
pred0 <- predict(m1, newdata = data_ctr, type = "response")
opt_treat <- ifelse(pred1 > pred0, 1, 0)
table(opt_treat)

# Pregnancy ended
data_trt <- data_ctr <- opt_mediation
data_trt$Group = "T"
data_ctr$Group = "C"
pred1 <- predict(m3, newdata = data_trt, type = "response")
pred0 <- predict(m3, newdata = data_ctr, type = "response")
opt_treat <- ifelse(pred1 > pred0, 1, 0)

table(opt_treat)
```

Write a simple piece of code which would return the optimal treatment decision for a given set of covariates. 

```{r warning = FALSE, message = FALSE}
# Birth weight
# Create variable for select covariates: a baseline fraction of bleeding when probed greater than/equal to 60 and age greater than/equal to 30
opt_g <- opt_mediation %>%
  mutate(Group = if_else(BL..BOP >= 60 & Age >= 30, 1, 0),
           Group = as.factor(Group)
         )

m1 <- glm(birthweight_imputed ~ Group*(Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I), data = opt_g, family = "gaussian")

data_trt <- data_ctr <- opt_g
data_trt$Group = "1"
data_ctr$Group = "0"
pred1 <- predict(m1, newdata = data_trt, type = "response")
pred0 <- predict(m1, newdata = data_ctr, type = "response")
opt_treat <- ifelse(pred1 > pred0, 1, 0)
table(opt_treat)

# Is treatment optimal for individuals with the given covariates?
isTRUE(mean(pred1) > mean(pred0))


# Pregnancy ended
m3 <- glm(Preg.ended...37.wk ~ Group*(Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I), data = opt_g, family = "binomial")

data_trt <- data_ctr <- opt_g
data_trt$Group = "1"
data_ctr$Group = "0"
pred1 <- predict(m3, newdata = data_trt, type = "response")
pred0 <- predict(m3, newdata = data_ctr, type = "response")
opt_treat <- ifelse(pred1 > pred0, 1, 0)
table(opt_treat)

# Is treatment optimal for individuals with the given covariates?
isTRUE(mean(pred0) > mean(pred1))

```

## Question 3. C-learning 

Taking a classification-learning (C-learning) perspective, estimate an optimal “decision list” treatment rule for each outcome in the OPT study. 

### Birthweight

- Step 1a: Fit a random forest to estimate $E(Y|A, X) = \mu(A, X)$

```{r warning = FALSE, message = FALSE}
data_trt_only <- opt_mediation[which(opt_mediation$Group == "T"), ]
data_ctr_only <- opt_mediation[which(opt_mediation$Group == "C"), ]
m1_treat <- ranger(birthweight_imputed ~ Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = data_trt_only, seed = 2123, probability = TRUE)
m1_ctr <- ranger(birthweight_imputed ~ Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = data_ctr_only, seed = 2123, probability = TRUE)
```

- Step 1b: Estimate the conditional average treatment effect as $Z_i = \hat{\mu}(1, X_i) - \hat{\mu}(0, X_i)$

```{r warning = FALSE, message = FALSE}
pred1 <- predict(m1_treat, data = data_trt)$predictions[, 2]
pred1[opt_mediation$Group == "T"] <- m1_treat$predictions[, 2]
pred0 <- predict(m1_ctr, data = data_ctr)$predictions[, 2]
pred0[opt_mediation$Group == "C"] <- m1_ctr$predictions[, 2]
opt_mediation$Z <- pred1 - pred0
```

Step 2:

```{r, fig.height=3, echo = TRUE}
opt_mediation$Z_label <- ifelse(opt_mediation$Z > 0, 1, 0)
opt_mediation$Z_weight <- abs(opt_mediation$Z)
fit <- rpart(Z_label ~ Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, 
	data = opt_mediation, weights = Z_weight,
	maxdepth = 3); rpart.plot(fit)
```


### Pregnancy ended

- Step 1a: Fit a random forest to estimate $E(Y|A, X) = \mu(A, X)$

```{r, echo = TRUE, warning = FALSE}
data_trt_only <- opt_mediation[which(opt_mediation$Group == "T"), ]
data_ctr_only <- opt_mediation[which(opt_mediation$Group == "C"), ]
m1_treat <- ranger(factor(Preg.ended...37.wk) ~ Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = data_trt_only, seed = 2123, probability = TRUE)
m1_ctr <- ranger(factor(Preg.ended...37.wk) ~ Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = data_ctr_only, seed = 2123, probability = TRUE)
```

- Step 1b: Estimate the conditional average treatment effect as $Z_i = \hat{\mu}(1, X_i) - \hat{\mu}(0, X_i)$

```{r warning = FALSE, message = FALSE}
pred1 <- predict(m1_treat, data = data_trt)$predictions[, 2]
pred1[opt_mediation$Group == "T"] <- m1_treat$predictions[, 2]
pred0 <- predict(m1_ctr, data = data_ctr)$predictions[, 2]
pred0[opt_mediation$Group == "C"] <- m1_ctr$predictions[, 2]
opt_mediation$Z <- pred1 - pred0
```

Step 2:

```{r, fig.height=3, echo = TRUE}
opt_mediation$Z_label <- ifelse(opt_mediation$Z > 0, 1, 0)
opt_mediation$Z_weight <- abs(opt_mediation$Z)
fit <- rpart(Z_label ~ Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed + drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, 
	data = opt_mediation, weights = Z_weight,
	maxdepth = 3); rpart.plot(fit)

```
