---
title: 'PUBH 7485 Homework 1'
output: 
  github_document:
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(tableone)
library(labelled)
library(rms)
library(boot)

options(scipen = 999)
options(digits = 3)
```

```{r warning = FALSE, message = FALSE}
# Read in data
load(here("data", "OPT_Study_PUBH7485_8485_2022.Rdata"))
```

### **Question 1:**

Summarize the differences between the treatment and control for the covariates.

```{r warning = FALSE, message = FALSE}
# Get variable names
dput(names(opt_causal))

# Vector of variables to summarize
myVars <- c(
  "Clinic", "Age", "Black", "White", "Nat.Am",
  "Asian", "Hisp", "Education", "Public.Asstce", "Hypertension",
  "Diabetes", "BL.Diab.Type", "BMI", "Use.Tob", "BL.Cig.Day", "Use.Alc",
  "BL.Drks.Day", "Drug.Add", "Prev.preg", "N.prev.preg", "Live.PTB",
  "Any.stillbirth", "Spont.ab", "Induced.ab", "Any.live.ptb.sb.sp.ab.in.ab",
  "N.living.kids", "N.qualifying.teeth", "BL.GE", "BL..BOP", "BL.PD.avg",
  "BL..PD.4", "BL..PD.5", "BL.CAL.avg", "BL..CAL.2", "BL..CAL.3",
  "BL.Calc.I", "BL.Pl.I", "Birth.outcome", "Preg.ended...37.wk",
  "GA.at.outcome", "Birthweight"
)

# Vector of categorical variables that need transformation
catVars <- c(
  "Clinic", "Group", "Black", "White", "Nat.Am",
  "Asian", "Hisp", "Education", "Public.Asstce", "Hypertension",
  "Diabetes", "BL.Diab.Type", "Use.Tob", "Use.Alc", "Drug.Add", "Prev.preg", "Live.PTB",
  "Any.stillbirth", "Spont.ab", "Induced.ab", "Any.live.ptb.sb.sp.ab.in.ab", "Birth.outcome",
  "Preg.ended...37.wk"
)

# Change variable labels
opt_causal <- set_variable_labels(opt_causal,
  Clinic = "Clinic location",
  Age = "Age (y)",
  Black = "Race: Black",
  White = "Race: White",
  Nat.Am = "Race: Native American",
  Asian = "Race: Asian",
  Hisp = "Ethnicity: Hispanic",
  Education = "Level of Education",
  Public.Asstce = "Received public assistance for delivery",
  Hypertension = "Chronic hypertension at baseline",
  Diabetes = "Diabetes at baseline",
  BL.Diab.Type = "Baseline diabetes type (for those with diabetes)",
  BMI = "Body mass index",
  Use.Tob = "History of tobacco use",
  BL.Cig.Day = "# cigarettes per day",
  Use.Alc = "History of alcohol use",
  BL.Drks.Day = "# drinks per day",
  Drug.Add = "History of drug addiction",
  Prev.preg = "Any previous pregnancy",
  N.prev.preg = "# previous pregnancies",
  Live.PTB = "Previous live preterm birth",
  Any.stillbirth = "Previous stillbirth",
  Spont.ab = "Previous spontaneous abortion",
  Induced.ab = "Previous induced abortion",
  Any.live.ptb.sb.sp.ab.in.ab = "Previous live pre-term birth, stillbirth,
or spontaneous/induced abortion",
  N.living.kids = "# of living children",
  N.qualifying.teeth = "# of qualifying teeth",
  BL.GE = "Whole-mouth average gingival index",
  BL..BOP = "Fraction of sites bleeding
on probing",
  BL.PD.avg = "Whole-mouth average pocket depth",
  BL..PD.4 = "Fraction of sites with pocket depth ≥ 4mm",
  BL..PD.5 = "Fraction of sites with pocket depth ≥ 5mm",
  BL.CAL.avg = "Whole-mouth average clinical attachment level",
  BL..CAL.2 = "Fraction of sites with clinical attachment level ≥ 2 mm",
  BL..CAL.3 = "Fraction of sites with clinical attachment level ≥ 3 mm",
  BL.Calc.I = "Whole-mouth average calculus index",
  BL.Pl.I = "Whole-mouth average plaque index",
  Birth.outcome = "Birth outcome",
  Preg.ended...37.wk = "Pregnancy ended before 37 weeks",
  GA.at.outcome = "Gestational age at outcome (d)",
  Birthweight = "Birth weight (g)"
)

# Create Table 1
tab1 <- CreateTableOne(vars = myVars, strata = "Group", data = opt_causal, factorVars = catVars, test = FALSE)

tab1Mat <- print(tab1, smd = TRUE, showAllLevels = TRUE, exact = "stage", quote = FALSE, noSpaces = TRUE, printToggle = FALSE, varLabels = TRUE)

## Save to CSV
write.csv(tab1Mat, file = "homework1_table1.csv")
```

### **Question 2:**

Find the unadjusted average treatment effect, standard error, and 95% confidence intervals for the outcomes of interest.

```{r warning = FALSE, message = FALSE}
# Outcome: Pregnancy ended < 37 weeks
## ATE
## Change group to numeric to calculate proportions
opt_causal_num <- opt_causal %>%
  mutate(Preg.ended...37.wk_num = fct_recode(Preg.ended...37.wk, "0" = "No ", "1" = "Yes")) %>%
  mutate(Preg.ended...37.wk_num = as.numeric(as.character(Preg.ended...37.wk_num))
  )

## Unadjusted ATE
p1 <- mean(opt_causal_num$Preg.ended...37.wk_num[opt_causal_num$Group == "T"])
p0 <- mean(opt_causal_num$Preg.ended...37.wk_num[opt_causal_num$Group == "C"])
p1 -  p0

## Standard error
n1 <- table(opt_causal_num$Group)[2]
n0 <- table(opt_causal_num$Group)[1]
SE <- sqrt(p1*(1 - p1)/n1 + p0*(1 - p0)/n0)
SE

## 95% CI
CI_unadj <- p1 - p0 + c(-1, 1)*qnorm(0.975)*SE
print(p1 - p0 + c(-1, 1)*qnorm(0.975)*SE, digits = 3)

# Outcome: Birth weight
## ATE & SE
lm_bw <- lm(Birthweight ~ Group, data = opt_causal)
summary(lm_bw)

## 95% CI
confint(lm_bw, level = 0.95)
```

## **Question 3:** 

Estimate the causal treatment effect using regression adjustment.

```{r warning = FALSE, message = FALSE}
# Function to impute missing values in continuous variables
imp_med <- function(x) {
  ifelse(is.na(x), median(x, na.rm = TRUE), x)
}

# Functions to impute missing values in categorical variables
calc_mode <- function(x){
    distinct_values <- unique(x)
    distinct_tabulate <- tabulate(match(x, distinct_values))
    distinct_values[which.max(distinct_tabulate)]
}

imp_cat <- function(x) {
  if_else(is.na(x), calc_mode(x), x)
}

# Create vector of continuous variables
cont_vars <- c(
  "Age","BMI", "BL.Cig.Day", "BL.Drks.Day", "N.prev.preg", "N.living.kids",
  "N.qualifying.teeth", "BL.GE", "BL..BOP", "BL.PD.avg",
  "BL..PD.4", "BL..PD.5", "BL.CAL.avg", "BL..CAL.2", "BL..CAL.3",
  "BL.Calc.I", "BL.Pl.I", "Birthweight"
)

# Create vector of categorical variables
cat_vars <- c(
  "Clinic","Black", "White", "Nat.Am", "Asian", "Hisp", "Education",
  "Public.Asstce", "Hypertension", "Diabetes", "BL.Diab.Type", "Use.Tob", 
  "Use.Alc", "Drug.Add", "Prev.preg", "Live.PTB", "Any.stillbirth", "Spont.ab", 
  "Induced.ab", "Any.live.ptb.sb.sp.ab.in.ab", "Preg.ended...37.wk"
)

# Impute median for continuous variables
for (i in cont_vars) {
  opt_causal[[i]] <- imp_med(opt_causal[[i]])
}
  
# Impute most common category for continuous variables
for (i in cat_vars) {
  opt_causal[[i]] <- imp_cat(opt_causal[[i]])
}

# Check for missing data
opt_causal %>% is.na() %>% colSums()
```

### **3.a:**

Regression models with main effect terms for the covariates and treatment.

```{r warning = FALSE, message = FALSE}
model1_preg <- glm(Preg.ended...37.wk ~ Group + Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = opt_causal, family = binomial(link = "logit"))

# Set up for ATE, SE, CI
data_trt <- data_ctr <- opt_causal
data_trt$Group = "T"
data_ctr$Group = "C"
pred1 <- predict(model1_preg, newdata = data_trt, type = "response")
pred0 <- predict(model1_preg, newdata = data_ctr, type = "response")
ATE <- mean(pred1 - pred0)
set.seed(100)
B <- 100
ATE_boot <- NULL
n <- nrow(opt_causal)

# Bootstrap
for(i in 1:B) {
  opt_causal_boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  model1_preg_boot <- glm(Preg.ended...37.wk ~ Group + Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = opt_causal_boot, family = binomial(link = "logit"))
  data_trt_boot <- opt_causal_boot
  data_trt_boot$Group = "T"
  data_ctr_boot <- opt_causal_boot
  data_ctr_boot$Group = "C"
  pred1_boot <- predict(model1_preg_boot, newdata = data_trt_boot, 
    type = "response")
  pred0_boot <- predict(model1_preg_boot, newdata = data_ctr_boot, 
    type = "response")
  ATE_boot <- c(ATE_boot, mean(pred1_boot - pred0_boot))
  }

# Function to print results
get_results <- function() {
  print("Average Treatment Effect")
  print(ATE, digits = 3)
  SE <- sd(ATE_boot) 
  print("Bootstrap SE")
  print(SE, digits = 3)
  print("Bootstrap Normal 95% CI")
  CI <- ATE + c(-1, 1)*qnorm(0.975)*SE
  print(ATE + c(-1, 1)*qnorm(0.975)*SE, digits = 3)
}

get_results()  
```

```{r warning = FALSE, message = FALSE}
# Create main effects model for Birth weight outcome
model1_bw <- glm(Birthweight ~ Group + Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = opt_causal)

# Set up for ATE, SE, CI
data_trt <- data_ctr <- opt_causal
data_trt$Group = "T"
data_ctr$Group = "C"
pred1 <- predict(model1_bw, newdata = data_trt, type = "response")
pred0 <- predict(model1_bw, newdata = data_ctr, type = "response")
ATE <- mean(pred1 - pred0)
set.seed(100)
B <- 100
ATE_boot <- NULL
n <- nrow(opt_causal)

# Bootstrap
for(i in 1:B) {
  opt_causal_boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  model1_bw_boot <- glm(Birthweight ~ Group + Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = opt_causal_boot)
  data_trt_boot <- opt_causal_boot
  data_trt_boot$Group = "T"
  data_ctr_boot <- opt_causal_boot
  data_ctr_boot$Group = "C"
  pred1_boot <- predict(model1_bw_boot, newdata = data_trt_boot, 
    type = "response")
  pred0_boot <- predict(model1_bw_boot, newdata = data_ctr_boot, 
    type = "response")
  ATE_boot <- c(ATE_boot, mean(pred1_boot - pred0_boot))
}

get_results()
```


###**3.b**	
Add in nonlinear terms for the continuous covariates but no interactions.

```{r warning = FALSE, message = FALSE}
# Pregnancy outcome
model2_preg <- glm(Preg.ended...37.wk ~ Group + Clinic + rcs(Age, 4) + Black + White + Nat.Am + Education + Public.Asstce + rcs(BMI, 4) + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + rcs(BL.GE, 4) + rcs(BL..BOP, 4) + rcs(BL.PD.avg, 4) + rcs(BL.CAL.avg, 4) + rcs(BL.Calc.I, 4) + rcs(BL.Pl.I, 4), data = opt_causal, family = binomial(link = "logit"))

# Function for bootstrap
ate.stat.rcs <- function(data, indices){
  data.boot <- data[indices,]
  
  m4.boot <- glm(Preg.ended...37.wk ~ Group + Clinic + rcs(Age, 4) + Black + White + Nat.Am + Education + Public.Asstce + rcs(BMI, 4) + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + rcs(BL.GE, 4) + rcs(BL..BOP, 4) + rcs(BL.PD.avg, 4) + rcs(BL.CAL.avg, 4) + rcs(BL.Calc.I, 4) + rcs(BL.Pl.I, 4), data = data.boot, family = binomial(link = "logit"))
  return(coef(m4.boot)[2])
}
# ATE
#p1 <- mean(opt_causal_num$Preg.ended...37.wk_num[opt_causal_num$Group == "T"])
#p0 <- mean(opt_causal_num$Preg.ended...37.wk_num[opt_causal_num$Group == "C"])
#ATE <- p1 -  p0

## Standard error
#n1 <- table(opt_causal$Group)[2]
#n0 <- table(opt_causal$Group)[1]
#SE <- sqrt(p1*(1 - p1)/n1 + p0*(1 - p0)/n0)

#LB <- ATE - qnorm(0.975)*SE
#UB <- ATE + qnorm(0.975)*SE
#estimator_name <- c("Unadjusted Analysis")


# Run bootstrap
set.seed(100)
boot.results <- boot(data = opt_causal, statistic = ate.stat.rcs, R = 1000, parallel = "multicore", ncpus = 8)
boot.results ; head(boot.results$t)

# Store results in matrix
ATE <- c(ATE, model2_preg$coefficients[2]); SE <- c(SE, sd(boot.results$t))
LB <- c(LB, sort(boot.results$t)[25]); UB <- c(UB, sort(boot.results$t)[975])
estimator_name <- c(estimator_name, "Main Effects, RCS")
print(paste0("ATE (SE) = ", round(ATE[3], 1), " (", round(SE[3], 1), ")"))
print(paste0("95% CI:", round(LB[3], 1), ", ", round(UB[3], 1)))

```


```{r warning = FALSE, message = FALSE}
# Birth weight outcome
model2_bw <- glm(Birthweight ~ Group + Clinic + rcs(Age, 4) + Black + White + Nat.Am + Education + Public.Asstce + rcs(BMI, 4) + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + rcs(BL.GE, 4) + rcs(BL..BOP, 4) + rcs(BL.PD.avg, 4) + rcs(BL.CAL.avg, 4) + rcs(BL.Calc.I, 4) + rcs(BL.Pl.I, 4), data = opt_causal)

summary(model2_bw)

# Function for bootstrap
ate.stat.rcs <- function(data, indices){
  data.boot <- data[indices,]
  
  m4.boot <- glm(Birthweight ~ Group + Clinic + rcs(Age, 4) + Black + White + Nat.Am + Education + Public.Asstce + rcs(BMI, 4) + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + rcs(BL.GE, 4) + rcs(BL..BOP, 4) + rcs(BL.PD.avg, 4) + rcs(BL.CAL.avg, 4) + rcs(BL.Calc.I, 4) + rcs(BL.Pl.I, 4), data = data.boot)
  return(coef(m4.boot)[2])
}

# Run bootstrap
set.seed(100)
boot.results <- boot(data = opt_causal, statistic = ate.stat.rcs, R = 1000, parallel = "multicore", ncpus = 8)
boot.results ; head(boot.results$t)


# Store results in matrix
ATE <- c(ATE, model2_bw$coefficients[2]); SE <- c(SE, sd(boot.results$t))
LB <- c(LB, sort(boot.results$t)[25]); UB <- c(UB, sort(boot.results$t)[975])
estimator_name <- c(estimator_name, "Main Effects, RCS")
print(paste0("ATE (SE) = ", round(ATE[3], 1), " (", round(SE[3], 1), ")"))
print(paste0("95% CI:", round(LB[3], 1), ", ", round(UB[3], 1)))

```


###**3.c**	
Regression model with interactions between the covariates and treatment; no nonlinear terms. 

```{r warning = FALSE, message = FALSE}
# Create interaction model for pregnancy outcome
model3_preg <- glm(Preg.ended...37.wk ~ Group*(Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I), data = opt_causal, family = binomial(link = "logit"))

# Set up for ATE, SE, CI
data_trt <- data_ctr <- opt_causal
data_trt$Group = "T"
data_ctr$Group = "C"
pred1 <- predict(model3_preg, newdata = data_trt, type = "response")
pred0 <- predict(model3_preg, newdata = data_ctr, type = "response")
ATE <- mean(pred1 - pred0)
set.seed(100)
B <- 100
ATE_boot <- NULL
n <- nrow(opt_causal)

# Bootstrap
for(i in 1:B) {
  opt_causal_boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  model3_preg_boot <- glm(Preg.ended...37.wk ~ Group + Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I, data = opt_causal_boot, family = binomial(link = "logit"))
  data_trt_boot <- opt_causal_boot
  data_trt_boot$Group = "T"
  data_ctr_boot <- opt_causal_boot
  data_ctr_boot$Group = "C"
  pred1_boot <- predict(model3_preg_boot, newdata = data_trt_boot, 
    type = "response")
  pred0_boot <- predict(model3_preg_boot, newdata = data_ctr_boot, 
    type = "response")
  ATE_boot <- c(ATE_boot, mean(pred1_boot - pred0_boot))
  }

get_results()

```


```{r warning = FALSE, message = FALSE}
# Create interaction model for birth weight outcome
model3_bw <- glm(Birthweight ~ Group*(Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I), data = opt_causal)

# Set up for ATE, SE, CI
data_trt <- data_ctr <- opt_causal
data_trt$Group = "T"
data_ctr$Group = "C"
pred1 <- predict(model3_bw, newdata = data_trt, type = "response")
pred0 <- predict(model3_bw, newdata = data_ctr, type = "response")
ATE <- mean(pred1 - pred0)
set.seed(100)
B <- 100
ATE_boot <- NULL
n <- nrow(opt_causal)

# Bootstrap
for(i in 1:B) {
  opt_causal_boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  model3_bw_boot <- glm(Birthweight ~ Group*(Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I), data = opt_causal_boot)
  data_trt_boot <- opt_causal_boot
  data_trt_boot$Group = "T"
  data_ctr_boot <- opt_causal_boot
  data_ctr_boot$Group = "C"
  pred1_boot <- predict(model3_bw_boot, newdata = data_trt_boot, 
    type = "response")
  pred0_boot <- predict(model3_bw_boot, newdata = data_ctr_boot, 
    type = "response")
  ATE_boot <- c(ATE_boot, mean(pred1_boot - pred0_boot))
}

get_results()
```
