---
title: 'PUBH 7485 Homework 2'
output: 
  github_document:
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(tableone)
library(labelled)
library(rms)
library(boot)
library(survey)
 
options(scipen = 999)
options(digits = 3)
```

```{r warning = FALSE, message = FALSE}
# Read in data
load(here("data", "OPT_Study_PUBH7485_8485_2022.Rdata"))
```


# Create Table 1
```{r warning = FALSE, message = FALSE}
# Vector of variables to summarize
myVars <- c(
  "Clinic", "Age", "Black", "White", "Nat.Am",
  "Asian", "Hisp", "Education", "Public.Asstce", "Hypertension",
  "Diabetes", "BL.Diab.Type", "BMI", "Use.Tob", "BL.Cig.Day", "Use.Alc",
  "BL.Drks.Day", "Drug.Add", "Prev.preg", "N.prev.preg", "Live.PTB",
  "Any.stillbirth", "Spont.ab", "Induced.ab", "Any.live.ptb.sb.sp.ab.in.ab",
  "N.living.kids", "N.qualifying.teeth", "BL.GE", "BL..BOP", "BL.PD.avg",
  "BL..PD.4", "BL..PD.5", "BL.CAL.avg", "BL..CAL.2", "BL..CAL.3",
  "BL.Calc.I", "BL.Pl.I", "Birth.outcome", "Preg.ended...37.wk",
  "GA.at.outcome", "Birthweight"
)

# Vector of categorical variables that need transformation
catVars <- c(
  "Clinic", "Group", "Black", "White", "Nat.Am",
  "Asian", "Hisp", "Education", "Public.Asstce", "Hypertension",
  "Diabetes", "BL.Diab.Type", "Use.Tob", "Use.Alc", "Drug.Add", "Prev.preg", "Live.PTB",
  "Any.stillbirth", "Spont.ab", "Induced.ab", "Any.live.ptb.sb.sp.ab.in.ab", "Birth.outcome",
  "Preg.ended...37.wk"
)

# Change variable labels
opt_causal <- set_variable_labels(opt_causal,
  Clinic = "Clinic location",
  Age = "Age (y)",
  Black = "Race: Black",
  White = "Race: White",
  Nat.Am = "Race: Native American",
  Asian = "Race: Asian",
  Hisp = "Ethnicity: Hispanic",
  Education = "Level of Education",
  Public.Asstce = "Received public assistance for delivery",
  Hypertension = "Chronic hypertension at baseline",
  Diabetes = "Diabetes at baseline",
  BL.Diab.Type = "Baseline diabetes type (for those with diabetes)",
  BMI = "Body mass index",
  Use.Tob = "History of tobacco use",
  BL.Cig.Day = "# cigarettes per day",
  Use.Alc = "History of alcohol use",
  BL.Drks.Day = "# drinks per day",
  Drug.Add = "History of drug addiction",
  Prev.preg = "Any previous pregnancy",
  N.prev.preg = "# previous pregnancies",
  Live.PTB = "Previous live preterm birth",
  Any.stillbirth = "Previous stillbirth",
  Spont.ab = "Previous spontaneous abortion",
  Induced.ab = "Previous induced abortion",
  Any.live.ptb.sb.sp.ab.in.ab = "Previous live pre-term birth, stillbirth,
or spontaneous/induced abortion",
  N.living.kids = "# of living children",
  N.qualifying.teeth = "# of qualifying teeth",
  BL.GE = "Whole-mouth average gingival index",
  BL..BOP = "Fraction of sites bleeding
on probing",
  BL.PD.avg = "Whole-mouth average pocket depth",
  BL..PD.4 = "Fraction of sites with pocket depth ≥ 4mm",
  BL..PD.5 = "Fraction of sites with pocket depth ≥ 5mm",
  BL.CAL.avg = "Whole-mouth average clinical attachment level",
  BL..CAL.2 = "Fraction of sites with clinical attachment level ≥ 2 mm",
  BL..CAL.3 = "Fraction of sites with clinical attachment level ≥ 3 mm",
  BL.Calc.I = "Whole-mouth average calculus index",
  BL.Pl.I = "Whole-mouth average plaque index",
  Birth.outcome = "Birth outcome",
  Preg.ended...37.wk = "Pregnancy ended before 37 weeks",
  GA.at.outcome = "Gestational age at outcome (d)",
  Birthweight = "Birth weight (g)"
)

# Create Table 1
tab1 <- CreateTableOne(vars = myVars, strata = "Group", data = opt_causal, factorVars = catVars, test = FALSE)

tab1Mat <- print(tab1, smd = TRUE, showAllLevels = TRUE, exact = "stage", quote = FALSE, noSpaces = TRUE, printToggle = FALSE, varLabels = TRUE)
```

### **Question 1:**

1.	Consider estimating the causal treatment effect using i) propensity score stratification and ii) inverse probability weighting (for the purposes of this assignment, please use the IPW2 estimator for the course notes). For each of these scenarios, please report the estimated average (causal) treatment effect, standard error, and 95% confidence interval in a table. Please have a separate table for each outcome and for each estimation approach (propensity score stratification and inverse probability weighting). Note that Birth.outcome and GA.at.outcome are additional outcome variables and PID is the patient identifier none of which should not be included in the regression models.

a)	Consider a logistic model for treatment assignment with just main effect terms for the covariates.
b)	Add in nonlinear terms for the continuous covariates but still no interactions.
c)	(BONUS) Consider some other model for treatment assignment such as a random forest.
d)	Consider a “final” propensity model to estimate the ATE which may include interactions and nonlinear terms. Variable selection may also be considered here as well. For this, in addition to providing the summary measures of the ATE described above, please describe how you arrived at your “final” propensity model.   
e)	Create a line plot of the weighted standardized mean difference using the propensity score model from (a) – (d) above as well as the unweighted standardized mean difference. Comment on the balance that weighting was able to achieve.   


```{r warning = FALSE, message = FALSE}
# Function to impute missing values in continuous variables
imp_med <- function(x) {
  ifelse(is.na(x), median(x, na.rm = TRUE), x)
}

# Functions to impute missing values in categorical variables
calc_mode <- function(x) {
  distinct_values <- unique(x)
  distinct_tabulate <- tabulate(match(x, distinct_values))
  distinct_values[which.max(distinct_tabulate)]
}

imp_cat <- function(x) {
  if_else(is.na(x), calc_mode(x), x)
}

# Create vector of continuous variables
cont_vars <- c(
  "Age", "BMI", "BL.Cig.Day", "BL.Drks.Day", "N.prev.preg", "N.living.kids",
  "N.qualifying.teeth", "BL.GE", "BL..BOP", "BL.PD.avg",
  "BL..PD.4", "BL..PD.5", "BL.CAL.avg", "BL..CAL.2", "BL..CAL.3",
  "BL.Calc.I", "BL.Pl.I", "Birthweight"
)

# Create vector of categorical variables
cat_vars <- c(
  "Clinic", "Black", "White", "Nat.Am", "Asian", "Hisp", "Education",
  "Public.Asstce", "Hypertension", "Diabetes", "BL.Diab.Type", "Use.Tob",
  "Use.Alc", "Drug.Add", "Prev.preg", "Live.PTB", "Any.stillbirth", "Spont.ab",
  "Induced.ab", "Any.live.ptb.sb.sp.ab.in.ab", "Preg.ended...37.wk"
)

# Impute median for continuous variables
for (i in cont_vars) {
  opt_causal[[i]] <- imp_med(opt_causal[[i]])
}

# Impute most common category for continuous variables
for (i in cat_vars) {
  opt_causal[[i]] <- imp_cat(opt_causal[[i]])
}

# Check for missing data
opt_causal %>%
  is.na() %>%
  colSums()
```

## Get propensity score
```{r warning = FALSE, message = FALSE}
# Create propensity models
# logistic main effects
p1 <- glm(Group ~ (Clinic + Age + Black + White + Nat.Am + Education + Public.Asstce + BMI + Hypertension + Diabetes + BL.Cig.Day + BL.Drks.Day + N.prev.preg + Any.live.ptb.sb.sp.ab.in.ab + N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I), data = opt_causal, family = "binomial")
round(summary(p1)$coefficients, digits = 3)

# Add non-linear terms
p2_nonlin <-  glm(Group ~ (Clinic + I(Age^2) + Black + White + Nat.Am + Education + Public.Asstce + I(BMI^2) + Hypertension + Diabetes + I(BL.Cig.Day^2) + I(BL.Drks.Day^2) + I(N.prev.preg^2) + Any.live.ptb.sb.sp.ab.in.ab + I(N.qualifying.teeth^2) + I(BL.GE^2) + I(BL..BOP^2) + I(BL.PD.avg^2) + I(BL.CAL.avg^2) + I(BL.Calc.I^2) + I(BL.Pl.I^2)), data = opt_causal, family = binomial)
round(summary(p2_nonlin)$coefficients, digits = 3)

## Obtain estimated propensity scores
opt_causal$ps <- predict(p1, type = "response")

```


## Obtain Estimated Propensity Scores and Divide Into Quintiles

```{r, echo = TRUE, cache = TRUE}
ps <- predict(p1, type = "response")
ps_quintile <- cut(ps, 
	breaks = c(0, quantile(ps, p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)
table(ps_quintile, opt_causal$Group)
```



## Check weighted differences in key variables between groups
```{r, echo = TRUE, warning=FALSE, results = "hide", size="tiny"} 
ps <- predict(p1, type = "response")
opt_causal$weight <- opt_causal$Group/ps + (1 - opt_causal$Group)/(1 - ps)
opt_svy <- svydesign(ids = ~ 1, data = opt_causal, weights = ~ weight)




## Weighted Differences in Key Variables Between Groups
ps <- predict(p1, type = "response")
imai$weight <- imai$PHN.C1/ps + (1-imai$PHN.C1)/(1-ps)
imaiSvy <- svydesign(ids = ~ 1, data = imai, weights = ~ weight)

tabWeighted <- svyCreateTableOne(vars = vars, strata = "PHN.C1F",
  data = imaiSvy, test = FALSE)
t2 <- print(tabWeighted, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)
```



```



## Propensity Scores Stratification

```{r, echo = TRUE, cache = TRUE}
n <- nrow(opt_causal)
nj <- table(ps_quintile)
te_quintile <- 
  tapply(opt_causal$Preg.ended[opt_causal$Group == "T"],
         ps_quintile[opt_causal$Group == "T"], mean) -
	tapply(opt_causal$Preg.ended[opt_causal$Group == "C"],
	       ps_quintile[opt_causal$Group == "C"], mean)
print(round(te_quintile, 3))
ATE_PSS <- sum(te_quintile *nj/n)
print(round(ATE_PSS, 3))

```


# 3.a: PSS regression adjustment
```{r, echo = TRUE, cache = TRUE}
# Create main effects model for preterm outcome
m1.ps <- glm(Preg.ended...37.wk ~ Group*rcs(ps, 5), data = opt_causal, family = "binomial")
round(summary(m1.ps)$coefficients, digits = 3)

# Set up for ATE, SE, CI
data_trt <- data_ctr <- opt_causal
data_trt$Group = "T"
data_ctr$Group = "C"
pred1.ps <- predict(m1.ps, newdata = data_trt, type = "response")
pred0.ps <- predict(m1.ps, newdata = data_ctr, type = "response")
ATE.ps <- mean(pred1.ps - pred0.ps)
print(ATE.ps, digits = 3)

# Bootstrap for PS Regression Adjustment
set.seed(100)
B <- 100
ATE.ps.boot <- NULL
n <- nrow(opt_causal)
for(i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  m1.ps.boot <- glm(Preg.ended...37.wk ~ Group*rcs(ps, 5), data = opt.boot, family = "binomial")
  data_trt.boot <- opt.boot
  data_trt.boot$Group = "T"
  data_ctr.boot <- opt.boot
  data_ctr.boot$Group = "C"
  pred1.ps.boot <- predict(m1.ps.boot, newdata = data_trt.boot, 
    type = "response")
  pred0.ps.boot <- predict(m1.ps.boot, newdata = data_ctr.boot, 
    type = "response")
  ATE.ps.boot <- c(ATE.ps.boot, mean(pred1.ps.boot - pred0.ps.boot))
}

# Function to print results
get_results <- function() {
  print("Average Treatment Effect")
  print(ATE.ps, digits = 3)
  SE.ps <- sd(ATE.ps.boot) 
  print("Bootstrap SE")
  print(SE.ps, digits = 3)
  print("Bootstrap Normal 95% CI")
  CI.ps <- ATE.ps + c(-1, 1)*qnorm(0.975)*SE.ps
  print(ATE.ps + c(-1, 1)*qnorm(0.975)*SE.ps, digits = 3)
}

get_results()
```

```{r warning = FALSE, message = FALSE}
# Create main effects model for Birth weight outcome
m2.ps <- glm(Birthweight ~ Group*rcs(ps, 5), data = opt_causal)
round(summary(m2.ps)$coefficients, digits = 3)

# Set up for ATE, SE, CI
data_trt <- data_ctr <- opt_causal
data_trt$Group <- "T"
data_ctr$Group <- "C"
pred1 <- predict(m2.ps, newdata = data_trt, type = "response")
pred0 <- predict(m2.ps, newdata = data_ctr, type = "response")
ATE.ps <- mean(pred1 - pred0)
print(ATE.ps, digits = 3)


# Bootstrap
set.seed(100)
B <- 100
ATE.ps.boot <- NULL
n <- nrow(opt_causal)
for(i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  m2.ps.boot <- glm(Birthweight ~ Group*rcs(ps, 5), data = opt.boot)
  data_trt.boot <- opt.boot
  data_trt.boot$Group = "T"
  data_ctr.boot <- opt.boot
  data_ctr.boot$Group = "C"
  pred1.ps.boot <- predict(m2.ps.boot, newdata = data_trt.boot, 
    type = "response")
  pred0.ps.boot <- predict(m2.ps.boot, newdata = data_ctr.boot, 
    type = "response")
  ATE.ps.boot <- c(ATE.ps.boot, mean(pred1.ps.boot - pred0.ps.boot))
}

get_results()
```

# 3.b: Add in nonlinear terms for the continuous covariates but still no interactions.

# Pregnancy outcome
model2_preg <- glm(Preg.ended...37.wk ~ Group + Clinic + I(Age^2) + Black + White + Nat.Am + Education + Public.Asstce + I(BMI^2) + Hypertension + Diabetes + I(BL.Cig.Day^2) + I(BL.Drks.Day^2) + I(N.prev.preg^2) + Any.live.ptb.sb.sp.ab.in.ab + I(N.qualifying.teeth^2) + I(BL.GE^2) + I(BL..BOP^2) + I(BL.PD.avg^2) + I(BL.CAL.avg^2) + I(BL.Calc.I^2) + I(BL.Pl.I^2), data = opt_causal, family = binomial(link = "logit"))

```{r, echo = TRUE, cache = TRUE}
# Create main effects model for preterm outcome
m1.ps <- glm(Preg.ended...37.wk ~ Group*rcs(ps, 5), data = opt_causal, family = "binomial")
round(summary(m1.ps)$coefficients, digits = 3)

# Set up for ATE, SE, CI
data_trt <- data_ctr <- opt_causal
data_trt$Group = "T"
data_ctr$Group = "C"
pred1.ps <- predict(m1.ps, newdata = data_trt, type = "response")
pred0.ps <- predict(m1.ps, newdata = data_ctr, type = "response")
ATE.ps <- mean(pred1.ps - pred0.ps)
print(ATE.ps, digits = 3)

# Bootstrap for PS Regression Adjustment
set.seed(100)
B <- 100
ATE.ps.boot <- NULL
n <- nrow(opt_causal)
for(i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  m1.ps.boot <- glm(Preg.ended...37.wk ~ Group*rcs(ps, 5), data = opt.boot, family = "binomial")
  data_trt.boot <- opt.boot
  data_trt.boot$Group = "T"
  data_ctr.boot <- opt.boot
  data_ctr.boot$Group = "C"
  pred1.ps.boot <- predict(m1.ps.boot, newdata = data_trt.boot, 
    type = "response")
  pred0.ps.boot <- predict(m1.ps.boot, newdata = data_ctr.boot, 
    type = "response")
  ATE.ps.boot <- c(ATE.ps.boot, mean(pred1.ps.boot - pred0.ps.boot))
}

# Function to print results
get_results <- function() {
  print("Average Treatment Effect")
  print(ATE.ps, digits = 3)
  SE.ps <- sd(ATE.ps.boot) 
  print("Bootstrap SE")
  print(SE.ps, digits = 3)
  print("Bootstrap Normal 95% CI")
  CI.ps <- ATE.ps + c(-1, 1)*qnorm(0.975)*SE.ps
  print(ATE.ps + c(-1, 1)*qnorm(0.975)*SE.ps, digits = 3)
}

get_results()
```

