---
title: 'PUBH 7485 Homework 3'
output: 
  github_document:
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(tableone)
library(labelled)
library(rms)
library(boot)
library(survey)
library(stats)
library(reshape2)
library(forestplot)
library(MatchIt)

options(scipen = 999)
options(digits = 3)
```

```{r warning = FALSE, message = FALSE}
# Read in data
load(here("data", "OPT_Study_PUBH7485_8485_2022.Rdata"))
```

## Optimal models
```{r warning = FALSE, message = FALSE}
# Impute continuous data
# Calculate medians
median_bmi <- median(opt_causal$BMI, na.rm = TRUE)
median_cigday <- median(opt_causal$BL.Cig.Day, na.rm = TRUE)
median_drinksday <- median(opt_causal$BL.Drks.Day, na.rm = TRUE)
median_prevpreg <- median(opt_causal$N.prev.preg, na.rm = TRUE)
median_livingkids <- median(opt_causal$N.living.kids, na.rm = TRUE)
median_birthweight <- median(opt_causal$Birthweight, na.rm = TRUE)
# Create new imputed columns
opt_causal$bmi_imputed <- ifelse(is.na(opt_causal$BMI), median_bmi, opt_causal$BMI)
opt_causal$cigday_imputed <- ifelse(is.na(opt_causal$BL.Cig.Day), median_cigday,
  opt_causal$BL.Cig.Day)
opt_causal$drinksday_imputed <- ifelse(is.na(opt_causal$BL.Drks.Day), 
  median_drinksday, opt_causal$BL.Drks.Day)
opt_causal$prevpreg_imputed <- ifelse(is.na(opt_causal$N.prev.preg), 
  median_prevpreg, opt_causal$N.prev.preg)
opt_causal$livingkids_imputed <- ifelse(is.na(opt_causal$N.living.kids), 
  median_livingkids, opt_causal$N.living.kids)
opt_causal$birthweight_imputed <- ifelse(is.na(opt_causal$Birthweight), 
  median_birthweight, opt_causal$Birthweight)

# Impute categorical data
# Consult Table One to find most common for each

# Hispanic: Yes
# Baseline Diabetes Type: Type II
# Tobacco Use: No
# Alcohol Use: No
# Drug addiction: No

# Create new imputed columns
opt_causal$hisp_imputed <- ifelse(is.na(opt_causal$Hisp), "Yes", opt_causal$Hisp)
opt_causal$diabtype_imputed <- ifelse(is.na(opt_causal$BL.Diab.Type), "Type II",
  opt_causal$BL.Diab.Type)
opt_causal$tobuse_imputed <- ifelse(is.na(opt_causal$Use.Tob), "No", opt_causal$Use.Tob)
opt_causal$alcuse_imputed <- ifelse(is.na(opt_causal$Use.Alc), "No", opt_causal$Use.Alc)
opt_causal$addiction_imputed <- ifelse(is.na(opt_causal$Drug.Add), "No", opt_causal$Drug.Add)

# Birth weight, linear regression
birth_weight_c <- lm(birthweight_imputed ~ Group*(Clinic + Age + I(Age^2) + Black + White +
    Nat.Am + Education + Public.Asstce + bmi_imputed + I(bmi_imputed^2) + Hypertension +
    Diabetes + cigday_imputed + I(cigday_imputed^2) + drinksday_imputed +
    I(drinksday_imputed^2) +
    prevpreg_imputed + I(prevpreg_imputed^2) + Any.live.ptb.sb.sp.ab.in.ab + 
    N.qualifying.teeth + I(N.qualifying.teeth^2) + BL.GE + I(BL.GE^2) + BL..BOP +
    I(BL..BOP^2) + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2) +
    BL.Calc.I + I(BL.Calc.I^2) + BL.Pl.I + I(BL.Pl.I^2)),
  data = opt_causal)
# Performed: step(birth_weight_c, direction = "both")
# Yields optimal model as:
birth_weight_c <- lm(birthweight_imputed ~ Group + Clinic + Age + Black + 
    White + Nat.Am + bmi_imputed + Hypertension + Diabetes + 
    cigday_imputed + I(cigday_imputed^2) + prevpreg_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.GE + I(BL.GE^2) + I(BL..BOP^2) + 
    BL.PD.avg + BL.CAL.avg + I(BL.Pl.I^2) + Group:Clinic + Group:Age + 
    Group:White + Group:Nat.Am + Group:bmi_imputed + Group:Hypertension + 
    Group:cigday_imputed + Group:I(cigday_imputed^2) + Group:Any.live.ptb.sb.sp.ab.in.ab + 
    Group:I(BL..BOP^2) + Group:BL.PD.avg + Group:BL.CAL.avg, data = opt_causal)

# End of pregnancy, logistic regression
preg_end_c <- glm(Preg.ended ~ Group*(Clinic + Age + Black + White + Nat.Am + Education +
    Public.Asstce + bmi_imputed + Hypertension + Diabetes + cigday_imputed +
    drinksday_imputed + prevpreg_imputed + Any.live.ptb.sb.sp.ab.in.ab + 
    N.qualifying.teeth + BL.GE + BL..BOP + BL.PD.avg + BL.CAL.avg + BL.Calc.I + BL.Pl.I),
  data = opt_causal, family = "binomial")
# Performed: step(preg_end_c, direction = "both")
# Yields optimal model as:
preg_end_c <- glm(Preg.ended ~ Group + Black + White + Nat.Am + Education + 
    Hypertension + Diabetes + Any.live.ptb.sb.sp.ab.in.ab + BL.GE + 
    BL.Pl.I + Group:Black + Group:White + Group:Nat.Am + Group:Education + 
    Group:Hypertension, family = "binomial", data = opt_causal)

# Group assignment, logistic regression

group_c <- glm(Group ~(Clinic + Age + I(Age^2) + Black + White +
    Nat.Am + Education + Public.Asstce + bmi_imputed + I(bmi_imputed^2) + Hypertension +
    Diabetes + cigday_imputed + I(cigday_imputed^2) + drinksday_imputed +
    I(drinksday_imputed^2) +
    prevpreg_imputed + I(prevpreg_imputed^2) + Any.live.ptb.sb.sp.ab.in.ab + 
    N.qualifying.teeth + I(N.qualifying.teeth^2) + BL.GE + I(BL.GE^2) + BL..BOP +
    I(BL..BOP^2) + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2) +
    BL.Calc.I + I(BL.Calc.I^2) + BL.Pl.I + I(BL.Pl.I^2)),
  data = opt_causal, family = "binomial")
# Performed: step(group_c, direction = "both")
# Yields optimal model as follows. I added in main effects for BL.PD.avg and BL.CAL.avg:
group_c <- glm(formula = Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), 
  family = "binomial", data = opt_causal)

```

# Question 1:

Using the “final” outcome and treatment models you developed in the previous assignment, find a “doubly robust” estimator of the ATE (as well as a standard error and confidence interval). How does the point estimate and standard error of this approach compare to the regression adjustment and IPW estimators? Please use both outcomes (whether or not pregnancy ended before 37 weeks and birthweight) considered in the first two assignments. If helpful, code will be provided to obtain a “final” outcome and treatment model that you may use. 


```{r warning = FALSE, message = FALSE}
p1 <- glm(Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_causal, family = binomial)
round(summary(p1)$coefficients, digits = 3)
```


## Form Pseudo-Outcomes for AIPW - Pregnancy ended
```{r warning = FALSE, message = FALSE}
# Recode group
opt_causal <- opt_causal %>%
  mutate(group_num = fct_recode(Group,
    "1" = "T",
    "0" = "C"
  )) %>%
  mutate(group_num = as.numeric(as.character(group_num)))

# Create model
m1 <- glm(Preg.ended ~ Group + Black + White + Nat.Am + Education + 
    Hypertension + Diabetes + Any.live.ptb.sb.sp.ab.in.ab + BL.GE + 
    BL.Pl.I + Group:Black + Group:White + Group:Nat.Am + Group:Education + 
    Group:Hypertension, data = opt_causal, family = "binomial")

# Set up ps and weights
ps <- predict(p1, type = "response")
w1 <- opt_causal$group_num / ps
w0 <- (1 - opt_causal$group_num) / (1 - ps)

# Get predictions
data_trt <- data_ctr <- opt_causal
data_trt$Group <- "T"
data_ctr$Group <- "C"
pred1 <- predict(m1, newdata = data_trt, type = "response")
pred0 <- predict(m1, newdata = data_ctr, type = "response")

PO1 <- opt_causal$Preg.ended * w1 - ((opt_causal$group_num - ps) / ps) * pred1
PO0 <- opt_causal$Preg.ended * w0 - ((1 - opt_causal$group_num - (1 - ps)) / (1 - ps)) * pred0

ATE_AIPW <- mean(PO1 - PO0, na.rm = TRUE)
ATE_AIPW
```


## AIPW Analysis - Pregnancy ended
```{r warning = FALSE, message = FALSE}
set.seed(2123)
B <- 100
ATE.boot <- NULL
n <- nrow(opt_causal)
for (i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  m1.boot <- glm(Preg.ended ~ Group + Black + White + Nat.Am + Education + 
    Hypertension + Diabetes + Any.live.ptb.sb.sp.ab.in.ab + BL.GE + 
    BL.Pl.I + Group:Black + Group:White + Group:Nat.Am + Group:Education + 
    Group:Hypertension,
    data = opt.boot,
    family = "binomial"
  )
  data_trt.boot <- opt.boot
  data_trt.boot$Group <- "T"
  data_ctr.boot <- opt.boot
  data_ctr.boot$Group <- "C"
  pred1.boot <- predict(m1.boot,
    newdata = data_trt.boot,
    type = "response"
  )
  pred0.boot <- predict(m1.boot,
    newdata = data_ctr.boot,
    type = "response"
  )
  p1.boot <- glm(Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2),
    data = opt.boot, family = "binomial"
  )
  ps.boot <- predict(p1.boot, type = "response")
  w1.boot <- opt.boot$group_num / ps.boot
  w0.boot <- (1 - opt.boot$group_num) / (1 - ps.boot)

  PO1.boot <- opt.boot$Preg.ended * w1.boot - ((opt.boot$group_num - ps.boot) / ps.boot) * pred1.boot
  PO0.boot <- opt.boot$Preg.ended * w0.boot - ((1 - opt.boot$group_num - (1 - ps.boot)) / (1 - ps.boot)) * pred0.boot

  ATE.boot <- c(
    ATE.boot,
    mean(PO1.boot - PO0.boot, na.rm = TRUE)
  )
}
print("Average Treatment Effect")
print(ATE_AIPW, digits = 3)
SE <- sd(ATE.boot)
print("Bootstrap SE")
print(SE, digits = 3)
print("Bootstrap Normal 95% CI")
CI_AIPW <- ATE_AIPW + c(-1, 1) * qnorm(0.975) * SE
print(ATE_AIPW + c(-1, 1) * qnorm(0.975) * SE, digits = 3)
```

## Form Pseudo-Outcomes for AIPW - Birth weight
```{r warning = FALSE, message = FALSE}
# Create model
m2 <- glm(birthweight_imputed ~ Group + Clinic + Age + Black + 
    White + Nat.Am + bmi_imputed + Hypertension + Diabetes + 
    cigday_imputed + I(cigday_imputed^2) + prevpreg_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.GE + I(BL.GE^2) + I(BL..BOP^2) + 
    BL.PD.avg + BL.CAL.avg + I(BL.Pl.I^2) + Group:Clinic + Group:Age + 
    Group:White + Group:Nat.Am + Group:bmi_imputed + Group:Hypertension + 
    Group:cigday_imputed + Group:I(cigday_imputed^2) + Group:Any.live.ptb.sb.sp.ab.in.ab + 
    Group:I(BL..BOP^2) + Group:BL.PD.avg + Group:BL.CAL.avg, data = opt_causal, family = "gaussian")

# Set up ps and weights
ps <- predict(p1, type = "response")
w1 <- opt_causal$group_num / ps
w0 <- (1 - opt_causal$group_num) / (1 - ps)

# Get predictions
data_trt <- data_ctr <- opt_causal
data_trt$Group <- "T"
data_ctr$Group <- "C"
pred1 <- predict(m2, newdata = data_trt, type = "response")
pred0 <- predict(m2, newdata = data_ctr, type = "response")

PO1 <- opt_causal$Preg.ended * w1 - ((opt_causal$group_num - ps) / ps) * pred1
PO0 <- opt_causal$Preg.ended * w0 - ((1 - opt_causal$group_num - (1 - ps)) / (1 - ps)) * pred0

ATE_AIPW <- mean(PO1 - PO0, na.rm = TRUE)
```


## AIPW Analysis - Birth weight
```{r warning = FALSE, message = FALSE}
set.seed(2123)
B <- 100
ATE.boot <- NULL
n <- nrow(opt_causal)
for (i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  m2.boot <- glm(birthweight_imputed ~ Group + Clinic + Age + Black + 
    White + Nat.Am + bmi_imputed + Hypertension + Diabetes + 
    cigday_imputed + I(cigday_imputed^2) + prevpreg_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.GE + I(BL.GE^2) + I(BL..BOP^2) + 
    BL.PD.avg + BL.CAL.avg + I(BL.Pl.I^2) + Group:Clinic + Group:Age + 
    Group:White + Group:Nat.Am + Group:bmi_imputed + Group:Hypertension + 
    Group:cigday_imputed + Group:I(cigday_imputed^2) + Group:Any.live.ptb.sb.sp.ab.in.ab + 
    Group:I(BL..BOP^2) + Group:BL.PD.avg + Group:BL.CAL.avg,
    data = opt.boot,
    family = "gaussian"
  )
  data_trt.boot <- opt.boot
  data_trt.boot$Group <- "T"
  data_ctr.boot <- opt.boot
  data_ctr.boot$Group <- "C"
  pred1.boot <- predict(m2.boot,
    newdata = data_trt.boot,
    type = "response"
  )
  pred0.boot <- predict(m2.boot,
    newdata = data_ctr.boot,
    type = "response"
  )
  p1.boot <- glm(Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2),
    data = opt.boot, family = "binomial"
  )
  ps.boot <- predict(p1.boot, type = "response")
  w1.boot <- opt.boot$group_num / ps.boot
  w0.boot <- (1 - opt.boot$group_num) / (1 - ps.boot)

  PO1.boot <- opt.boot$Birthweight * w1.boot - ((opt.boot$group_num - ps.boot) / ps.boot) * pred1.boot
  PO0.boot <- opt.boot$Birthweight * w0.boot - ((1 - opt.boot$group_num - (1 - ps.boot)) / (1 - ps.boot)) * pred0.boot

  ATE.boot <- c(
    ATE.boot,
    mean(PO1.boot - PO0.boot, na.rm = TRUE)
  )
}
print("Average Treatment Effect")
print(ATE_AIPW, digits = 3)
SE <- sd(ATE.boot)
print("Bootstrap SE")
print(SE, digits = 3)
print("Bootstrap Normal 95% CI")
CI_AIPW <- ATE_AIPW + c(-1, 1) * qnorm(0.975) * SE
print(ATE_AIPW + c(-1, 1) * qnorm(0.975) * SE, digits = 3)
```

# Question 2:

Consider estimating the average treatment effect among the treated. Again, please consider both outcomes (whether or not pregnancy ended before 37 weeks and birthweight) considered in the first assignment. Using the “final” outcome and treatment models from the first assignment, find the ATT using
a.	Regression adjustment with covariates
b.	Regression adjustment with propensity score
c.	Propensity score stratification
d.	Inverse probability weighting
e.	Propensity score matching (consider both 1:1 and 2:1 matching)

```{r warning = FALSE, message = FALSE}
# Regression adjustment with covariates

# Regression adjustment: Get predicted value for each individual in the TREATED GROUP assuming that they are (a) in the treatment group and (b) in the control group and then take the difference in the mean predicted value to get estimate of ATT.

# Pregnancy ended
data_trt <- data_ctr <- opt_causal[opt_causal$Group == "T", ]
data_trt$Group <- "T"
data_ctr$Group <- "C"
pred1 <- predict(m1, newdata = data_trt, type = "response")
pred0 <- predict(m1, newdata = data_ctr, type = "response")
ATT <- mean(pred1 - pred0)
print(ATT, digits = 3)

ATT_boot <- NULL
for (i in 1:B) {
  opt_causal_boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  model1_preg_boot <- glm(Preg.ended ~ Group + Black + White + Nat.Am + Education + 
    Hypertension + Diabetes + Any.live.ptb.sb.sp.ab.in.ab + BL.GE + 
    BL.Pl.I + Group:Black + Group:White + Group:Nat.Am + Group:Education + 
    Group:Hypertension, data = opt_causal_boot, family = binomial(link = "logit"))
  data_trt_boot <- opt_causal_boot
  data_trt_boot$Group <- "T"
  data_ctr_boot <- opt_causal_boot
  data_ctr_boot$Group <- "C"
  pred1_boot <- predict(model1_preg_boot,
    newdata = data_trt_boot,
    type = "response"
  )
  pred0_boot <- predict(model1_preg_boot,
    newdata = data_ctr_boot,
    type = "response"
  )
  ATT_boot <- c(ATT_boot, mean(pred1_boot - pred0_boot))
}

SE <- sd(ATT_boot)
print("Bootstrap SE")
print(SE, digits = 3)
print("Bootstrap Normal 95% CI")
CI <- ATT + c(-1, 1) * qnorm(0.975) * SE
print(ATT + c(-1, 1) * qnorm(0.975) * SE, digits = 3)

# Birth weight
data_trt <- data_ctr <- opt_causal[opt_causal$Group == "T", ]
data_trt$Group <- "T"
data_ctr$Group <- "C"
pred1 <- predict(m2, newdata = data_trt, type = "response")
pred0 <- predict(m2, newdata = data_ctr, type = "response")
ATT <- mean(pred1 - pred0)
print(ATT, digits = 3)

ATT_boot <- NULL
for (i in 1:B) {
  opt_causal_boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  model2_bw_boot <- glm(birthweight_imputed ~ Group + Clinic + Age + Black + 
    White + Nat.Am + bmi_imputed + Hypertension + Diabetes + 
    cigday_imputed + I(cigday_imputed^2) + prevpreg_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.GE + I(BL.GE^2) + I(BL..BOP^2) + 
    BL.PD.avg + BL.CAL.avg + I(BL.Pl.I^2) + Group:Clinic + Group:Age + 
    Group:White + Group:Nat.Am + Group:bmi_imputed + Group:Hypertension + 
    Group:cigday_imputed + Group:I(cigday_imputed^2) + Group:Any.live.ptb.sb.sp.ab.in.ab + 
    Group:I(BL..BOP^2) + Group:BL.PD.avg + Group:BL.CAL.avg, data = opt_causal_boot,
    family = gaussian())
  data_trt_boot <- opt_causal_boot
  data_trt_boot$Group <- "T"
  data_ctr_boot <- opt_causal_boot
  data_ctr_boot$Group <- "C"
  pred1_boot <- predict(model2_bw_boot,
    newdata = data_trt_boot,
    type = "response"
  )
  pred0_boot <- predict(model2_bw_boot,
    newdata = data_ctr_boot,
    type = "response"
  )
  ATT_boot <- c(ATT_boot, mean(pred1_boot - pred0_boot))
}

SE <- sd(ATT_boot)
print("Bootstrap SE")
print(SE, digits = 3)
print("Bootstrap Normal 95% CI")
CI <- ATT + c(-1, 1) * qnorm(0.975) * SE
print(ATT + c(-1, 1) * qnorm(0.975) * SE, digits = 3)

```


```{r warning = FALSE, message = FALSE}
# Regression adjustment with propensity score

# Pregnancy ended
opt_causal$ps <- predict(p1, type = "response")
m1.ps <- glm(Preg.ended ~ Group * rcs(ps, 5), data = opt_causal, family = "binomial")
data_trt_alt <- data_ctr_alt <- opt_causal
data_trt_alt$Group <- "T"
data_ctr_alt$Group <- "C"
pred1.ps <- predict(m1.ps, newdata = data_trt_alt, type = "response")
pred0.ps <- predict(m1.ps, newdata = data_ctr_alt, type = "response")
ATT.ps <- mean(pred1.ps[opt_causal$Group == "T"] - pred0.ps[opt_causal$Group == "C"])
print(ATT.ps, digits = 3)

set.seed(2321)
B <- 100
ATT.ps.boot <- NULL

n <- nrow(opt_causal)
for(i in 1:B) {
opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
m1.ps.boot <- glm(Preg.ended ~ group_num*rcs(ps, 5), data = opt.boot, family = "binomial")
data_trt_alt.boot <- data_ctr_alt.boot <- opt.boot
data_trt_alt.boot$group_num = 1
data_ctr_alt.boot$group_num = 0
pred1.ps.boot <- predict(m1.ps.boot, newdata = data_trt_alt.boot, type = "response")
pred0.ps.boot <- predict(m1.ps.boot, newdata = data_ctr_alt.boot, type = "response")
ATT.ps.boot <- c(ATT.ps.boot, mean(pred1.ps.boot[opt.boot$group_num == 1] - pred0.ps.boot[opt.boot$group_num == 1]))
}

SE <- sd(ATT.ps.boot)
CI.ps <- ATT.ps + c(-1, 1)*qnorm(0.975)*sd(ATT.ps.boot)
SE
CI.ps


# Birth weight
opt_causal$ps <- predict(p1, type = "response")
m2.ps <- glm(birthweight_imputed ~ Group * rcs(ps, 5), data = opt_causal, family = "gaussian")
data_trt_alt <- data_ctr_alt <- opt_causal
data_trt_alt$Group <- "T"
data_ctr_alt$Group <- "C"
pred1.ps <- predict(m2.ps, newdata = data_trt_alt, type = "response")
pred0.ps <- predict(m2.ps, newdata = data_ctr_alt, type = "response")
ATT.ps <- mean(pred1.ps[opt_causal$Group == "T"] - pred0.ps[opt_causal$Group == "C"])
print(ATT.ps, digits = 3)


set.seed(2321)
B <- 100
ATT.ps.boot <- NULL
n <- nrow(opt_causal)
for(i in 1:B) {
opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
m1.ps.boot <- glm(birthweight_imputed ~ group_num*rcs(ps, 5), data = opt.boot, family = "gaussian")
data_trt_alt.boot <- data_ctr_alt.boot <- opt.boot
data_trt_alt.boot$group_num = 1
data_ctr_alt.boot$group_num = 0
pred1.ps.boot <- predict(m1.ps.boot, newdata = data_trt_alt.boot, type = "response")
pred0.ps.boot <- predict(m1.ps.boot, newdata = data_ctr_alt.boot, type = "response")
ATT.ps.boot <- c(ATT.ps.boot, mean(pred1.ps.boot[opt.boot$group_num == 1] - pred0.ps.boot[opt.boot$group_num == 1]))
}

SE <- sd(ATT.ps.boot)
CI.ps <- ATT.ps + c(-1, 1)*qnorm(0.975)*sd(ATT.ps.boot)
SE
CI.ps
```

```{r warning = FALSE, message = FALSE}
# PSS
# Propensity score stratification: Form strata using quintiles of ps in TREATED GROUP. Weight by nAj/nA where nA is number in treated group and nAj is number of treated in stratum j.

# Pregnancy ended
ps_quintile <- cut(ps,
  breaks = c(0, quantile(ps[opt_causal$Group == "T"], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5
)
nA <- nrow(opt_causal[opt_causal$Group == "T", ])
nAj <- table(ps_quintile[opt_causal$Group == "T"])
te_quintile <- tapply(opt_causal$Preg.ended[opt_causal$Group == "T"], ps_quintile[opt_causal$Group == "T"], mean) -
  tapply(opt_causal$Preg.ended[opt_causal$Group == "C"], ps_quintile[opt_causal$Group == "C"], mean)
ATT_PSS <- sum(te_quintile * nAj / nA)
print(round(ATT_PSS, 3))

# Note: the outcome needs to be coded numerically

set.seed(2321)
B <- 100
ATT_PSS.boot <- NULL
n <- nrow(opt_causal)
for (i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  p1.boot <- glm(Preg.ended ~ Group + Black + White + Nat.Am + Education + 
    Hypertension + Diabetes + Any.live.ptb.sb.sp.ab.in.ab + BL.GE + 
    BL.Pl.I + Group:Black + Group:White + Group:Nat.Am + Group:Education + 
    Group:Hypertension, data = opt.boot, family = "binomial")
  ps.boot <- predict(p1.boot, type = "response", newdata = opt.boot)
  ps_quintile.boot <- cut(ps.boot, 
	breaks = c(0, quantile(ps.boot[opt.boot$group_num == 1], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5)
  nA.boot <- nrow(opt.boot[opt.boot$group_num == 1, ])
  nAj.boot <- table(ps_quintile.boot[opt.boot$group_num == 1])
  te_quintile.boot <- tapply(opt.boot$Preg.ended[opt.boot$group_num == 1], 
  	ps_quintile.boot[opt.boot$group_num == 1], mean) -
	tapply(opt.boot$Preg.ended[opt.boot$group_num == 0], ps_quintile.boot[opt.boot$group_num == 0], mean) # this line results in NA
  ATT_PSS.boot <- c(ATT_PSS.boot, sum(te_quintile.boot *nAj.boot/nA.boot))
}

print(ATT_PSS, digits = 3)
SE_PSS <- sd(ATT_PSS.boot)
print("Bootstrap SE")
print(SE_PSS, digits = 3)
print("Bootstrap Normal 95% CI")
print(ATT_PSS + c(-1, 1) * qnorm(0.975) * SE_PSS, digits = 3)
CI_PSS <- ATT_PSS + c(-1, 1) * qnorm(0.975) * SE_PSS


# Birth weight
ps_quintile <- cut(ps,
  breaks = c(0, quantile(ps[opt_causal$Group == "T"], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5
)
nA <- nrow(opt_causal[opt_causal$Group == "T", ])
nAj <- table(ps_quintile[opt_causal$Group == "T"])
te_quintile <- tapply(opt_causal$birthweight_imputed[opt_causal$Group == "T"], ps_quintile[opt_causal$Group == "T"], mean) -
  tapply(opt_causal$birthweight_imputed[opt_causal$Group == "C"], ps_quintile[opt_causal$Group == "C"], mean)
ATT_PSS <- sum(te_quintile * nAj / nA)
print(round(ATT_PSS, 3))

set.seed(2321)
B <- 100
ATT_PSS.boot <- NULL
n <- nrow(opt_causal)
for (i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  p1.boot <- glm(birthweight_imputed ~ Group + Clinic + Age + Black + 
    White + Nat.Am + bmi_imputed + Hypertension + Diabetes + 
    cigday_imputed + I(cigday_imputed^2) + prevpreg_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.GE + I(BL.GE^2) + I(BL..BOP^2) + 
    BL.PD.avg + BL.CAL.avg + I(BL.Pl.I^2) + Group:Clinic + Group:Age + 
    Group:White + Group:Nat.Am + Group:bmi_imputed + Group:Hypertension + 
    Group:cigday_imputed + Group:I(cigday_imputed^2) + Group:Any.live.ptb.sb.sp.ab.in.ab + 
    Group:I(BL..BOP^2) + Group:BL.PD.avg + Group:BL.CAL.avg, data = opt.boot, family = "gaussian")
  ps.boot <- predict(p1.boot, type = "response", newdata = opt.boot)
  ps_quintile.boot <- cut(ps.boot, 
	breaks = c(0, quantile(ps.boot[opt.boot$group_num == 1], p = c(0.2, 0.4, 0.6, 0.8)), 1), labels = 1:5) # getting NAs in this part
  nA.boot <- nrow(opt.boot[opt.boot$group_num == 1, ])
  nAj.boot <- table(ps_quintile.boot[opt.boot$group_num == 1])
  te_quintile.boot <- tapply(opt.boot$birthweight_imputed[opt.boot$group_num == 1], 
  	ps_quintile.boot[opt.boot$group_num == 1], mean) -
	tapply(opt.boot$birthweight_imputed[opt.boot$group_num == 0], ps_quintile.boot[opt.boot$group_num == 0], mean)
  ATT_PSS.boot <- c(ATT_PSS.boot, sum(te_quintile.boot *nAj.boot/nA.boot))
}

print(ATT_PSS, digits = 3)
SE_PSS <- sd(ATT_PSS.boot)
print("Bootstrap SE")
print(SE_PSS, digits = 3)
print("Bootstrap Normal 95% CI")
print(ATT_PSS + c(-1, 1) * qnorm(0.975) * SE_PSS, digits = 3)
CI_PSS <- ATT_PSS + c(-1, 1) * qnorm(0.975) * SE_PSS
```


```{r warning = FALSE, message = FALSE}
# IPW2
# IPW: Weights for treated subjects are 1 and for control are π(Xi )(1 − Ai )/{1 − π(Xi )} (must use IPW2 here)

# IPW2 Adjustment
# Pregnancy ended
ps <- predict(p1, type = "response")
w1 <- opt_causal$group_num
w0 <- (1 - opt_causal$group_num) / (1 - ps) * ps
ATT_IPW2 <- weighted.mean(opt_causal$Preg.ended, w = w1) -
  weighted.mean(opt_causal$Preg.ended, w = w0)
print(ATT_IPW2, digits = 3)

# Note: the outcome needs to be coded numerically

set.seed(2123)
B <- 100
ATT_IPW2.boot <- NULL
n <- nrow(opt_causal)
for (i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  p1.boot <- glm(Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2),
    data = opt.boot, family = "binomial"
  )
  ps.boot <- predict(p1.boot, type = "response")
  w1.boot <- opt.boot$group_num / ps.boot
  w0.boot <- (1 - opt.boot$group_num) / (1 - ps.boot)
  ATT_IPW2.boot <- c(ATT_IPW2.boot,
    weighted.mean(opt.boot$Preg.ended, w = w1.boot) - weighted.mean(opt.boot$Preg.ended, w = w0.boot)
  )
}

SE_IPW2 <- sd(ATT_IPW2.boot)
print("Bootstrap SE")
print(SE_IPW2, digits = 3)
print("Bootstrap Normal 95% CI")
print(ATT_IPW2 + c(-1, 1) * qnorm(0.975) * SE_IPW2, digits = 3)
CI_IPW2 <- ATT_IPW2 + c(-1, 1) * qnorm(0.975) * SE_IPW2


# Birth weight
ps <- predict(p1, type = "response")
w1 <- opt_causal$group_num
w0 <- (1 - opt_causal$group_num) / (1 - ps) * ps
ATT_IPW2 <- weighted.mean(opt_causal$birthweight_imputed, w = w1) -
  weighted.mean(opt_causal$birthweight_imputed, w = w0)
print(ATT_IPW2, digits = 3)

set.seed(2123)
B <- 100
ATT_IPW2.boot <- NULL
n <- nrow(opt_causal)
for (i in 1:B) {
  opt.boot <- opt_causal[sample(1:n, n, replace = TRUE), ]
  p1.boot <- glm(Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2),
    data = opt.boot, family = "binomial"
  )
  ps.boot <- predict(p1.boot, type = "response")
  w1.boot <- opt.boot$group_num / ps.boot
  w0.boot <- (1 - opt.boot$group_num) / (1 - ps.boot)
  ATT_IPW2.boot <- c(ATT_IPW2.boot,
    weighted.mean(opt.boot$birthweight_imputed, w = w1.boot) - weighted.mean(opt.boot$birthweight_imputed, w = w0.boot)
  )
}

SE_IPW2 <- sd(ATT_IPW2.boot)
print("Bootstrap SE")
print(SE_IPW2, digits = 3)
print("Bootstrap Normal 95% CI")
print(ATT_IPW2 + c(-1, 1) * qnorm(0.975) * SE_IPW2, digits = 3)
CI_IPW2 <- ATT_IPW2 + c(-1, 1) * qnorm(0.975) * SE_IPW2

```

```{r warning = FALSE, message = FALSE}
# Propensity score matching (consider both 1:1 and 2:1 matching)
## 1:1 Matching - Preg ended

rownames(opt_causal) <- 1:nrow(opt_causal)
mod_match_1 <- matchit(Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), distance = "logit", method = "nearest", data = opt_causal, ratio = 1)
# print(tail(mod_match_1$match.matrix))
summary(mod_match_1)

## Results 1:1 Matching
p1_m <- mean(opt_causal$Preg.ended[opt_causal$Group == "T"])
p0_m <- mean(opt_causal$Preg.ended[as.numeric(mod_match_1$match.matrix)])
n1 <- table(opt_causal$Group)[2]
n0 <- length(as.numeric(mod_match_1$match.matrix))
n0 <- table(opt_causal$Group)[1]
print("1:1 Match ATT")
ATT1match <- p1_m - p0_m
print(p1_m - p0_m, digits = 3)
print("Standard Error")
SE <- sqrt(p1_m * (p1_m) / n1 + p0_m * (1 - p0_m) / n0)
print(SE, digits = 3)
print("95% CI")
CI1match <- p1_m - p0_m + c(-1, 1) * qnorm(0.975) * SE
print(p1_m - p0_m + c(-1, 1) * qnorm(0.975) * SE, digits = 3)

## 2:1 Matching - Preg ended
rownames(opt_causal) <- 1:nrow(opt_causal)
mod_match_2 <- matchit(Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), distance = "logit", method = "nearest", data = opt_causal, ratio = 2)
# print(tail(mod_match_2$match.matrix))
summary(mod_match_2)

## Results 2:1 Matching
p1_m <- mean(opt_causal$Preg.ended[opt_causal$Group == "T"])
p0_m <- mean(opt_causal$Preg.ended[as.numeric(mod_match_2$match.matrix)])
n1 <- table(opt_causal$Group)[2]
n0 <- length(as.numeric(mod_match_2$match.matrix))
n0 <- table(opt_causal$Group)[1]
print("2:1 Match ATT")
ATT1match <- p1_m - p0_m
print(p1_m - p0_m, digits = 3)
print("Standard Error")
SE <- sqrt(p1_m * (p1_m) / n1 + p0_m * (1 - p0_m) / n0)
print(SE, digits = 3)
print("95% CI")
CI1match <- p1_m - p0_m + c(-1, 1) * qnorm(0.975) * SE
print(p1_m - p0_m + c(-1, 1) * qnorm(0.975) * SE, digits = 3)
```


```{r warning = FALSE, message = FALSE}
# Propensity score matching
## 1:1 Matching - Birthweight
## Results 1:1 Matching
p1_m <- mean(opt_causal$birthweight_imputed[opt_causal$Group == "T"])
p0_m <- mean(opt_causal$birthweight_imputed[as.numeric(mod_match_1$match.matrix)])
n1 <- table(opt_causal$Group)[2]
n0 <- length(as.numeric(mod_match_1$match.matrix))
#n0 <- table(opt_causal$Group)[1]
print("1:1 Match ATT")
ATT1match <- p1_m - p0_m
print(p1_m - p0_m, digits = 3)
print("Standard Error")
SE <- sqrt(p1_m*(1-p1_m)/n1 + p0_m*(1-p0_m)/n0) # negative
print(SE, digits = 3)
print("95% CI")
CI1match <- p1_m - p0_m + c(-1, 1) * qnorm(0.975) * SE
print(p1_m - p0_m + c(-1, 1) * qnorm(0.975) * SE, digits = 3)

## 2:1 Matching - Birthweight
## Results 2:1 Matching
p1_m <- mean(opt_causal$birthweight_imputed[opt_causal$Group == "T"])
p0_m <- mean(opt_causal$birthweight_imputed[as.numeric(mod_match_2$match.matrix)])
n1 <- table(opt_causal$Group)[2]
n0 <- length(as.numeric(mod_match_2$match.matrix))
#n0 <- table(opt_causal$Group)[1]
print("1:1 Match ATT")
ATT1match <- p1_m - p0_m
print(p1_m - p0_m, digits = 3)
print("Standard Error")
SE <- sqrt(p1_m * (1-p1_m)/n1 + p0_m * (1 - p0_m) / n0) # negative
print(SE, digits = 3)
print("95% CI")
CI1match <- p1_m - p0_m + c(-1, 1) * qnorm(0.975) * SE
print(p1_m - p0_m + c(-1, 1) * qnorm(0.975) * SE, digits = 3)

```

## Question 3: 

For the propensity score matching use tabular and graphical diagnostics to assess the quality of the matching. Please include a line plot of the SMD for each covariates (include the unmatched cohort and using IPW as well). 

## Graphically Assess Matching
```{r, echo = TRUE}
plot(mod_match_1, type = "hist")
plot(mod_match_2, type = "hist")
```

## Graphically Assess Matching
```{r, echo = TRUE}
plot(mod_match_1, type = "QQ", which.xs = c("Age"))
plot(mod_match_2, type = "QQ", which.xs = c("Age"))

```

## Plot of SMD comparing matching

```{r warning = FALSE, message = FALSE}
# Vector of variables to summarize
vars <- c(
  "Clinic", "Age", "Black", "White", "Nat.Am", "Education", "Public.Asstce", "Hypertension",
  "Diabetes", "bmi_imputed", "cigday_imputed", "drinksday_imputed", "prevpreg_imputed", 
  "Any.live.ptb.sb.sp.ab.in.ab", "N.qualifying.teeth", "BL.GE", "BL..BOP", "BL.PD.avg",
  "BL.CAL.avg", "BL.Calc.I", "BL.Pl.I"
)

# Vector of categorical variables that need transformation
catVars <- c(
  "Clinic", "Group", "Black", "White", "Nat.Am", "Education", "Public.Asstce", "Hypertension",
  "Diabetes", "Any.live.ptb.sb.sp.ab.in.ab"
)

# Change variable labels
opt_causal <- set_variable_labels(opt_causal,
  Clinic = "Clinic location",
  Age = "Age (y)",
  Black = "Race: Black",
  White = "Race: White",
  Nat.Am = "Race: Native American",
  Education = "Level of Education",
  Public.Asstce = "Received public assistance for delivery",
  Hypertension = "Chronic hypertension at baseline",
  Diabetes = "Diabetes at baseline",
  bmi_imputed = "Body mass index",
  cigday_imputed = "# cigarettes per day",
  drinksday_imputed = "# drinks per day",
  prevpreg_imputed = "# previous pregnancies",
  Any.live.ptb.sb.sp.ab.in.ab = "Previous live pre-term birth, stillbirth,
or spontaneous/induced abortion",
  N.qualifying.teeth = "# of qualifying teeth",
  BL.GE = "Whole-mouth average gingival index",
  BL..BOP = "Fraction of sites bleeding
on probing",
  BL.PD.avg = "Whole-mouth average pocket depth",
  BL.CAL.avg = "Whole-mouth average clinical attachment level",
  BL.Calc.I = "Whole-mouth average calculus index",
  BL.Pl.I = "Whole-mouth average plaque index",
)

# Unmatched SMD
tabUnmatched <- CreateTableOne(vars = vars, strata = "Group", data = opt_causal, factorVars = catVars, test = FALSE)

tab1 <- print(tabUnmatched, smd = TRUE, showAllLevels = TRUE, exact = "stage", quote = FALSE, noSpaces = TRUE, printToggle = FALSE, varLabels = TRUE)
write.csv(tab1, file = "homework3_table1.csv")

# Find weighted differences (IPW)
p1 <- glm(Group ~ White + Nat.Am + Diabetes + cigday_imputed + 
    Any.live.ptb.sb.sp.ab.in.ab + BL.PD.avg + I(BL.PD.avg^2) + BL.CAL.avg + I(BL.CAL.avg^2), data = opt_causal, family = binomial)
ps_a <- predict(p1, type = "response")
opt_causal$weight <- opt_causal$group_num / ps_a + (1 - opt_causal$group_num) / (1 - ps_a)
opt_svy <- svydesign(ids = ~1, data = opt_causal, weights = ~weight)
tabWeighted <- svyCreateTableOne(vars = vars, strata = "Group", data = opt_svy, test = FALSE)

t2 <- print(tabWeighted, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)
print(tabWeighted, smd = TRUE, showAllLevels = TRUE, exact = "stage", quote = FALSE, noSpaces = TRUE, printToggle = FALSE, varLabels = TRUE)
write.csv(t2, file = "homework3_table2.csv")

# Find SMD 1:1: ps matching
opt_match <- data.frame(rbind(opt_causal[opt_causal$group_num == 1, ],
	opt_causal[as.numeric(mod_match_1$match.matrix), ]))
tabmatched1.1 <- CreateTableOne(vars = vars, strata = "Group",
	data = opt_match, test = FALSE)
t3 <- print(tabmatched1.1, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)
print(tabmatched1.1, smd = TRUE, showAllLevels = TRUE, exact = "stage", quote = FALSE, noSpaces = TRUE, printToggle = FALSE, varLabels = TRUE)
write.csv(t3, file = "homework3_table3.csv")


# Find SMD 2:1: ps matching
opt_match <- data.frame(rbind(opt_causal[opt_causal$group_num == 1, ],
	opt_causal[as.numeric(mod_match_2$match.matrix), ]))
tabmatched2.1 <- CreateTableOne(vars = vars, strata = "Group",
	data = opt_match, test = FALSE)
t4 <- print(tabmatched2.1, smd = TRUE, showAllLevels = TRUE, varLabels = TRUE)
print(tabmatched2.1, smd = TRUE, showAllLevels = TRUE, exact = "stage", quote = FALSE, noSpaces = TRUE, printToggle = FALSE, varLabels = TRUE)
write.csv(t4, file = "homework3_table4.csv")


dataPlot <- data.frame(
  variable = rownames(ExtractSmd(tabUnmatched)),
  Unweighted = as.numeric(ExtractSmd(tabUnmatched)),
  IPW2 = as.numeric(ExtractSmd(tabWeighted)),
  Match_1 = as.numeric(ExtractSmd(tabmatched1.1)),
  Match_2 = as.numeric(ExtractSmd(tabmatched2.1))
)
dataPlot <- dplyr::filter(
  dataPlot,
  variable != "Preg.ended" &
    variable != "Preg.ended...37.wk" &
    variable != "Birthweight"
)

## Create long-format data for ggplot2
dataPlotMelt <- melt(
  data = dataPlot,
  id.vars = c("variable"),
  variable.name = "Method",
  value.name = "SMD"
)

## Order variable names by magnitude of SMD
varNames <- as.character(dataPlot$variable)[order(dataPlot$Unweighted)]

## Order factor levels in the same order
dataPlotMelt$variable <- factor(dataPlotMelt$variable,
  levels = varNames
)

## Plot using ggplot2
ggplot(
  data = dataPlotMelt,
  mapping = aes(x = variable, y = SMD, group = Method, color = Method)
) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.1, color = "black", size = 0.1) +
  coord_flip() +
  labs(title = "Figure 1. Standardized mean difference various analysis approaches\n") +
  theme_minimal() +
  theme(
    legend.key = element_blank(),
    plot.title = element_text(size = 12, hjust = 0.5)
  )
```